---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

If zip file exists, unzip and read file. Display first few data points to confirm.
```{r loadData}
if (file.exists("activity.zip")) {
    raw_data <- read.csv(unz("activity.zip", "activity.csv"))
    raw_data$full_date <- strptime(raw_data$date, "%Y-%m-%d")
    head(raw_data)
} else {
    stop("you need to put your activity.zip file right here next to this script...")
}
```


## What is mean total number of steps taken per day?
Remove NAs, and then calculate sum of steps per day.
```{r stepsPerDay}
filled_data <- raw_data[complete.cases(raw_data$steps),]
steps_per_day <- setNames(aggregate(filled_data$steps ~ filled_data$date, factor(filled_data$date), sum), c('date', 'steps'))
```


Calculate mean of all steps per day.
```{r meanSteps}
mean(steps_per_day$steps)
```
Also calculate mean per day of steps.
```{r meanDaySteps}
mean_steps_per_day <- setNames(aggregate(filled_data$steps ~ filled_data$date, factor(filled_data$date), mean), c('date', 'mean_steps'))
head(mean_steps_per_day)
```

Calculate median steps per day.
```{r medianSteps}
median(steps_per_day$steps)
```
Also calculate median per day of steps.
```{r medianDaySteps}
median_steps_per_day <- setNames(aggregate(filled_data$steps ~ filled_data$date, factor(filled_data$date), median), c('date', 'median_steps'))
head(median_steps_per_day)
```

Plot histogram of steps per day.
```{r plotHist}
hist(steps_per_day$steps, main = "Histogram of Steps per Day", xlab = "Steps per Day")
```

## What is the average daily activity pattern?

Find steps per interval
```{r stepsPerInterval}
steps_per_interval <- setNames(aggregate(filled_data$steps ~ filled_data$interval, factor(filled_data$interval), mean), c('interval', 'mean_steps'))
```
Plot average steps per interval
```{r plotInterval}
plot(steps_per_interval, type="l")
```
Return interval of maximum number of steps.
```{r maxInterval}
steps_per_interval[steps_per_interval$mean_steps == max(steps_per_interval$mean_steps), "interval"]
```
## Imputing missing values
Calculate number of rows with NA value for steps.
```{r missingRows}
length(which(is.na(raw_data$steps)))
```
Replace NAs with mean value for that interval
```{r replaceNAs}
replace_data <- merge(raw_data, steps_per_interval, all = TRUE)
replace_data$steps[is.na(replace_data$steps)] <- replace_data$mean_steps[is.na(replace_data$steps)]
```
Compare steps per day mean and median
```{r compareNA}
new_steps_per_day <- setNames(aggregate(replace_data$steps ~ replace_data$date, factor(replace_data$date), sum), c('date', 'steps'))
mean(new_steps_per_day$steps)
median(new_steps_per_day$steps)
```
Plot histogram of steps per day with filled-in na values.
```{r plotNAHist}
hist(new_steps_per_day$steps, main = "Histogram of Steps per Day (with filled-in NA values)", xlab = "Steps per Day")
```
## Are there differences in activity patterns between weekdays and weekends?
Create new column for weekend/weekday designation.
```{r weekends}
weekend = c("Saturday", "Sunday")
replace_data$weekday <- ifelse(weekdays(replace_data$full_date) %in% weekend, "weekend", "weekday")
```
